Combining checkpointing and data compression\\ for high frequency seismic imaging

### 2018-10-03 paper/paper.tex
revised Introduction

\begin{IEEEkeywords}
component, formatting, style, styling, insert
\end{IEEEkeywords}

\section{Introduction}
The class of adjoint-based optimization problems has a very typical data access pattern where a
simulation is run ``forward" in simulation time, which produces data that is to be used by a
subsequent adjoint pass which is run backward in simulation time. Figure \ref{fig:dataflow}
shows the typical data access pattern. This class consists of many different problems of interest in
science and engineering. One problem in this class is full-waveform inversion (FWI) where physical
experiments are carried out by propagating seismic waves through the earth's subsurface and the
solution to the optimization problem gives the most likely structure of the earth's subsurface. Figure
\ref{fig:offshore_survey} shows the physical setup of the experiment that produces the data that this problem operates
on. \cite{plessix2006review} The data access pattern from Figure \ref{fig:dataflow} applies here as well, since every step during the
adjoint propagation requires data from the corresponding step from the forward propagation. In
particular for FWI, this data can reach tens of terabytes in size, which is beyond the memory
available on most hardware. While this paper is written to be focussed on FWI as the target problem, 
the analysis can be applied to any problem with a similar data-access pattern.

@@

Various strategies are routinely employed to reduce this memory requirement, for example 
Domain decomposition where the domain of the simulation is distributed over
multiple nodes so as to fit within the memory available on each node. This introduces the overhead
of network communication and might not efficiently use the computational resources on each node.\cite{virieux2009seismic}
Another common strategy, which is specific to the domain of FWI, is where only the boundaries of
the domain are stored at each timestep and the rest of the wavefield is reconstructed from the
boundaries when required \cite{clapp2009reverse}\cite{yang2014rtm}. Another commonly used strategy that is applied across
a much wider domain of problems is checkpointing. In this strategy, a subset of the data from the
forward pass is stored (and the rest discarded). The discarded data is recomputed when required
by rerunning the forward pass. Starting from Revolve \cite{griewank2000algorithm}, which is a provably optimal
scheduling algorithm to implement this strategy, various studies have used and extended this
technique to various scenarios \cite{wang2009minimal} \cite{aupy2016optimal} \cite{schanen2016asynchronous}.

@@

Another strategy to reduce the memory footprint of scientific applications is
data compression. Well-known general purpose data compression algorithms like
Zlib (which is a part of gzip) \cite{deutsch1996zlib}, and compression for video and image
data such as JPEG-2000 \cite{skodras2001jpeg}have been used for scientific data. More
recently, special purpose compression algorithms for floating-point scientific data have been
developed, such as ZFP or SZ \cite{lindstrom2014fixed} \cite{di2018efficient} 
\cite{Kaklamanis:2012aa}.

Lossless algorithms guarantee that the exact original data can be
recovered during decompression, whereas lossy algorithms introduce an error, but
often guarantee that the error does not exceed certain absolute or relative
error metrics. Typically, lossy compression is more effective in reducing the
data size, or in other words, offer a greater compression ratio. Most popular
compression packages offer various settings that allow a tradeoff between
compression ratio, accuracy, and compression and decompression time. In this
work we use the ZFP package to perform lossy compresssion.

It is worth noting that another popular data reduction strategy is to typecast
values into a lower precision format, for example, from double precision to
single precision TODO cite. This can be seen as a computationally cheap lossy
compression algorithm with a compression ratio of $2$.

Perhaps counterintuitively, compression can not only reduce the memory
footprint, but also speed up an application. Previous work has observed that the
compression and decompression time can be less than the time saved from the
reduction in data that needs to be communicated across MPI nodes or between a
GPU and a host computer. There has been some work on streaming compression
of floating-point data for this specific intent. \cite{gpu-compression}

One possible way of using compression to speed up full-waveform inversion or any adjoint-based
optimization is by simply compressing the field data before storing it and decompressing it when it
is required again. If the memory available is sufficient to store the complete forward data after
compression, the use of compression effectively replaces the use of revolve-based checkpointing.
This scenario has been discussed in different studies before - they look at Revolve as an alternative
to compression. This approach was seen for a study of adjoint-based methods on the Navier-Stokes
equation \cite{cyr2015towards}. This was also studied in the context of seismic imaging and FWI
to develop tailored compression methods for the wavefield data that is typically seen in this context - 
including a compression method based on DCT \cite{dalmau2014lossy} and one based on adaptive
coarsening \cite{boehm2016wavefield}. This paper focusses on a broader region, of which this scenario
is just one part. In particular, we discuss the scenarios where the forward field can not fit in memory even
after compression, where compression and Revolve would be used in conjunction.

Although most of the compression we discuss here introduces some error which may or may not affect
the final result of the computation, we try to avoid discussing the error in this paper and focus on the 
performance impact of compression instead.

>>>>

\begin{IEEEkeywords}
Checkpointing, compression, adjoints, inversion
\end{IEEEkeywords}

\section{Introduction}
Adjoint-based optimization problems typically consist of a simulation that is
run forward in simulation time, producing data that is used in reverse order by
a subsequent adjoint computation that is run backwards in simulation time.
Figure~\ref{fig:dataflow} shows the resulting data flow. Many important
applications in science and engineering have this structure, including
full-waveform inversion (FWI).

In FWI, the propagation of seismic waves through the earth's subsurface is
simulated and compared with data from real-world measurements. The model of the
subsurface is iteratively improved by minimizing the misfit between simulation and
measurement in an adjoint optimization problem.
Figure~\ref{fig:offshore_survey} shows the setup of the real-world experiment that
produces the measurement data~\cite{plessix2006review}. Storing all intermediate
results as required for the data flow shown in Figure~\ref{fig:dataflow} is not
feasible, since this would require tens of terabytes of memory, which is not
commonly available on the machines used to solve these problems. To solve this
problem, a number of strategies are used that are discussed in the following
paragraphs. While this paper uses FWI as a test case, most of these ideas,
including the one presented in this work, are also applicable to other adjoint
optimization problems.

@@

Domain decomposition is often used not only to distribute the computational
workload across more processors, but also to utilise the large amount of memory
available in distributed systems. While this strategy is very powerful, the
number of compute nodes and therefore the amount of memory that can be used
efficiently is limited, for example by communication overheads that start to
dominate as the domain is split into increasingly small
pieces~\cite{virieux2009seismic}.
Another common strategy in FWI is to only store values at the boundaries of the
domain at each timestep, and reconstruct the rest of the wavefield when
required~\cite{clapp2009reverse,yang2014rtm}.

Checkpointing is yet another strategy to reduce the memory overhead. Only a
subset of the program states during the forward pass is stored (and the rest
discarded).  The discarded data is recomputed when needed by restarting the
forward pass from the last available stored state. The question which states
should be stored and which states should be recomputed to minimize the total
amount of recomputation work has been answerd under certain assumptions with the
Revolve algorithm~\cite{griewank2000algorithm}.  Other authors have subsequently
developed extensions to Revolve that are optimal under less restrictive
assumptions~\cite{wang2009minimal,aupy2016optimal,schanen2016asynchronous}.

@@

Data compression is also increasingly used to reduce the memory footprint of
scientific applications. General purpose data compression algorithms like Zlib
(which is a part of gzip)~\cite{deutsch1996zlib}, and compression algorithms for
video and image data such as JPEG-2000~\cite{skodras2001jpeg} have been
presented in previous work. More recently, special purpose compression
algorithms for floating-point scientific data have been developed, such as ZFP
or SZ~\cite{Kaklamanis:2012aa,lindstrom2014fixed,di2018efficient}.

Lossless algorithms guarantee that the exact original data can be recovered
during decompression, whereas lossy algorithms introduce an error, but often
guarantee that the error does not exceed certain absolute or relative error
metrics. Typically, lossy compression is more effective in reducing the data
size. Most popular compression packages offer various settings that allow a
tradeoff between compression ratio, accuracy, and compression and decompression
time. In this work we use the ZFP package to perform lossy compresssion.

It is worth noting that another data reduction strategy is to typecast values
into a lower precision format, for example, from double precision to single
precision. This can be seen as a computationally cheap lossy compression
algorithm with a compression ratio of $2$.

Perhaps counterintuitively, compression can not only reduce the memory
footprint, but also speed up an application. Previous work has observed that the
compression and decompression time can be less than the time saved from the
reduction in data that needs to be communicated across MPI nodes or between a
GPU and a host computer~\cite{gpu-compression}.

Another way of using compression to speed up FWI or other adjoint-based methods
is to use it \emph{instead} of checkpointing. If the compression ratio is
sufficient to fit the entire data in memory, checkpointing is no longer
necessary. Previous work has discussed this situation in the context of
computational fluid dynamics~\cite{cyr2015towards} and FWI using compression
method specifically designed for
wavefields~\cite{dalmau2014lossy,boehm2016wavefield}.

In this paper, we extend the previous studies by \emph{combining} checkpointing
and compression. An obvious use case would be one where the data does not fit
in the available memory even after compression. This could happen either for
very large adjoint problems, or for problems where the required accuracy limits
the achievable compression ratios.

Compared to using only checkpointing without compression, our combined method
can often achieve better performance. This is achieved by reducing the size of
stored program states, thereby allowing more program states to be stored during
the forward computation. This in turn reduces the amount of recomputations that
needs to be performed during the adjoint computation. On the other hand, the
compression and decompression itself takes time.  We thereforen provide a
comprehensive performance model that predicts whether or not the combined method
is beneficial. We present various scenarios that use a TODO describe Richter and
TODO describe JLSE/SKX node system, error tolerances for the lossy ZFP
compression algorithm ranging from TODO to TODO, and different types of forward
and adjoint computations that vary in their ratio between compute cost and state
size.

Our aim is to make this discussion independent of the hardware, application and
compression algorithm. We therefore do not discuss whether or not the accuracy
of the decompressed data is sufficient for our application, or whether
other algorithms might achieve a better accuracy/compression tradeoff.
